Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-10 ms.qmd
summary

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Scheiner2020]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010; @Zheng2023]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020]. The implications of changes in variance could therefore have wide-reaching implications for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

@@

Understanding how the acclimation capacity and the opportunity for selection will change across populations and species is important for predicting the impacts of climate change. Here, we show that the opportunity for selection is expected to decrease for wild ectotherms under climate change. This is because the acute and acclimation responses of wild ectotherms are expected to be lower in warmer environments.

>>>>

Reversible forms of phenotypic plasticty, such as acclimitisation and behavioural plasticity, are expected to provide greater potential to buffer populations from climate impacts as responses can be fine-tuned to environments [assuming the costs of plasticty are low – @Scheiner2020; @Dewitt1998]. Such responses are driven by changes in underlying physiology that can respond rapidly to the environment. However, studies have primarily been focused on mean physiological responses, paying little attention to changes in physiological variability [@Seebacher2015-wi; @Havird2020-mj]. For example, most work has focused on the degree to which mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022], neglecting how intrapopulation variability might also be impacted. Understanding how variability in physiological rates -- traits thought to be closely closely linked to fitness -- are affected by climate change has important implications for understanding a populations capacity for physiological trait evolution because lack of physiological variation is expected to play a role in limiting responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variability may also imply greater niche breadth and inform on the extent to which populations are able to buffer against environmental change [for example, through 'portfolio effects' see @Schindler2010; @Zheng2023]. Decreases in phenotypic variance may also suggest strong stabilising selection or constraints on performance, depending on the trait [@Scheiner2020]. The implications of changes in variance could therefore have wide-reaching implications for understanding the capacity for populations to adapt to novel environments [@Scheiner2020].

@@

Understanding how the acclimation capacity and the opportunity for selection will change across populations and species is important for predicting the impacts of climate change [@Chevin2017; @Chevin2010]. Here, we show that the opportunity for selection is expected to decrease for wild ectotherms, but particularly so for terrestrial ectotherms (~28%). Decreases in variance in physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the opportunity for selection for marine and terrestrial ectotherms under climate change (between 4-13% reduction in total variance).
